---
title: "Practical Machine Learning Course Project"
author: "Renan Almeida Baqui"
date: "02/11/2021"
output: html_document
html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

**Practical Machine Learning by Johns Hopkins University**

**Course Project**

**Background**

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now
possible to collect a large amount of data about personal activity
relatively inexpensively. These type of devices are part of the
quantified self movement -- a group of enthusiasts who take measurements
about themselves regularly to improve their health, to find patterns in
their behavior, or because they are tech geeks. One thing that people
regularly do is quantify how much of a particular activity they do, but
they rarely quantify how well they do it. In this project, your goal
will be to use data from accelerometers on the belt, forearm, arm, and
dumbell of 6 participants. They were asked to perform barbell lifts
correctly and incorrectly in 5 different ways. More information is
available from the website here:
<http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight
Lifting Exercise Dataset).

**Data**

The training data for this project are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

The data for this project come from this source:
<http://groupware.les.inf.puc-rio.br/har>. If you use the document you
create for this class for any purpose please cite them as they have been
very generous in allowing their data to be used for this kind of
assignment.

**What you should submit**

The goal of your project is to predict the manner in which they did the
exercise. This is the "classe" variable in the training set. You may use
any of the other variables to predict with. You should create a report
describing how you built your model, how you used cross validation, what
you think the expected out of sample error is, and why you made the
choices you did. You will also use your prediction model to predict 20
different test cases.

**The Project**

**Libraries**

Loading the R packages and libraries

```{r}
library(caret)
library(knitr)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)

set.seed(13343)
```

**Downloading the Data**

Extracting the two data sets from the provided URLs

```{r}
trainingData <- read.csv('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv',header=T)
testingData <- read.csv('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv',header=T)
```

Creating two subsets within the training data set: *train* (75%) and
*cross validation* (25%)

```{r}
 
inTrain  <- createDataPartition(trainingData$classe, p=0.75, list=FALSE)
TrainSet <- trainingData[inTrain, ]
CrossValSet  <- trainingData[-inTrain, ]
dim(TrainSet)
dim(CrossValSet)
```

**Pre-processing**

Verifying and exploring the training data set

When looking at the training data set table we can see that there
are a lot of NA variables that need pre processing and some variables are only for identification purposes

Removing nearly zero variance (NZV) variables from both sets

```{r}
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
CrossValSet  <- CrossValSet[, -NZV]
dim(TrainSet)
dim(CrossValSet)
```

Removing NA variables (more than 70% of NA in the variable)

```{r}
AllNA    <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.7
TrainSet <- TrainSet[, AllNA==FALSE]
CrossValSet  <- CrossValSet[, AllNA==FALSE]
dim(TrainSet)
dim(CrossValSet)
```

Removing unnecessary qualitative variables (columns 1 to 7)

```{r}
TrainSet <- TrainSet[, -(1:7)]
CrossValSet  <- CrossValSet[, -(1:7)]
dim(TrainSet)
dim(CrossValSet)
```

**Training the Machine Learning Models**

*Decision Tree Model*

Model fit

```{r}
set.seed(13343)
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
fancyRpartPlot(modFitDecTree)
```

Prediction and Confusion Matrix on Cross Validation data set

```{r}
predictDecTree <- predict(modFitDecTree, newdata=CrossValSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, as.factor(CrossValSet$classe))
confMatDecTree
```

Plot and model accuracy

```{r}
plot(confMatDecTree$table, col = confMatDecTree$byClass, 
     main = paste("Decision Tree Accuracy =",
                  round(confMatDecTree$overall['Accuracy'], 3)))
```

Expected out of sample error for Decision Tree: 1 - 0.695 = 0.305

*Random Forest*

Model fit

```{r}
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf",
                          trControl=controlRF)
modFitRandForest$finalModel
```

Prediction and Confusion Matrix on Cross Validation data set

```{r}
predictRandForest <- predict(modFitRandForest, newdata=CrossValSet)
confMatRandForest <- confusionMatrix(predictRandForest, as.factor(CrossValSet$classe))
confMatRandForest
```

Plot and model accuracy

```{r}
plot(confMatRandForest$table, col = confMatRandForest$byClass, 
     main = paste("Random Forest Accuracy =",
                  round(confMatRandForest$overall['Accuracy'], 3)))
```

Expected out of sample error for Random Forest: 1 - 0.992 = 0.008

*Generalized Boosted Model (GBM)*

Model fit

```{r}
set.seed(13343)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM  <- train(classe ~ ., data=TrainSet, method = "gbm",
                    trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel
```

Prediction and Confusion Matrix on Cross Validation data set

```{r}
predictGBM <- predict(modFitGBM, newdata=CrossValSet)
confMatGBM <- confusionMatrix(predictGBM, as.factor(CrossValSet$classe))
confMatGBM
```

Plot and model accuracy

```{r}
plot(confMatGBM$table, col = confMatGBM$byClass, 
     main = paste("GBM Accuracy =", round(confMatGBM$overall['Accuracy'], 3)))
```

Expected out of sample error for the Generalized Boosted Model: 1 -
0.959 = 0.041

**Prediction on Test Cases**

Since the *Random Forest Model* provided the best accuracy (and less
expected out of sample error) among the three tested models earlier,
this is the one that is going to be chosen for my final prediction

Final prediction on testing data (Course Project Prediction Quiz)

```{r}
predictQuiz <- predict(modFitRandForest, newdata=testingData)
predictQuiz
```
